---
format: gfm
default-image-extension: ""
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  fig.width = 9,
  fig.height = 3.2,
  dpi = 200,
  out.width = "100%"
)
```

# foresty

Forest plots and subgroup effects, taken from models you have already fitted.

An interaction p-value tells you that subgroups depart from a common effect.
It does not tell you which subgroup carries the effect, in which direction, or
how far apart they are. A small p-value from a large study can sit above two
subgroup estimates that are practically the same, and a large one can sit above
two that are not. The p-value on its own is not enough to decide what to
report.

`foresty` puts the two side by side. Give it a fitted model, the name of the
exposure and the name of an effect modifier, and it adds the interaction term,
estimates the exposure effect within every level of the modifier, tests the
interaction, and draws the result.

## What it is for: two-way interactions

**`foresty` is a package for two-way interactions, and only for those.** One
exposure crossed with one effect modifier, per figure: `exposure = "no2"` and
`interaction = "sex"` gives the effect of NO<sub>2</sub> within each level of
sex, and the joint test of the NO<sub>2</sub> by sex terms beside it. That
two-way interaction is the one a paper reports in a subgroup panel and the one
this package is built to make readable.

What follows from that:

* **One modifier per call.** Two of them are two calls, put together with
  `foresty_combine()` as two blocks of one figure — sex in one block, maternal
  smoking in the next — never crossed with each other. A three-way interaction
  is not fitted, not estimated and not tested.
* **The test is of that pair.** The p-value beside a block is the joint test
  that every exposure-by-modifier coefficient is zero, and nothing else in the
  model is being tested by it.
* **Higher-order terms already in your model are left alone.** A model that
  happens to carry a three-way interaction is used as it was fitted; the
  subgroup estimates are then reported at the first complete observation's
  values of the other interacting variables, and `foresty` warns that it has
  done so. Read those figures knowing what they hold.

## Installation

```{r}
#| eval: false
# install.packages("remotes")
remotes::install_github("AkiShiroshita/foresty")
```

## Use

Three functions. `foresty_main()` draws the effect of one exposure from each of
several models, one row apiece:

```{r}
#| label: main
library(foresty)

fit_no2 <- glm(asthma ~ no2 + sex + maternal_smoking + maternal_age,
               family = binomial, data = foresty_cohort)
fit_bc <- glm(asthma ~ black_carbon + sex + maternal_smoking + maternal_age,
              family = binomial, data = foresty_cohort)

foresty_main(
  list(fit_no2, fit_bc),
  exposure = c("no2", "black_carbon"),
  labels = c(no2 = "NO2", black_carbon = "Black carbon")
)
```

The figure says which effect it is showing and that it came out of a model
with no interaction in it, which is what separates it from the next one.

`foresty_interaction()` is the one the package exists for:

```{r}
#| label: by-sex
by_sex <- foresty_interaction(fit_no2, exposure = "no2", interaction = "sex")

by_sex             # draws the figure
```

```{r}
#| label: by-sex-summary
summary(by_sex)    # the numbers behind it
```

The interaction is tested by the likelihood ratio test against the same model
without it. `test = "wald"` takes the joint Wald test over its terms instead,
and `test = "both"` reports each in a column of its own — worth doing to show
that the p-value does not turn on which test was taken. The two usually agree;
where they do not, which is a small study, a rare outcome or a sparse subgroup,
the likelihood ratio test is the one to trust, which is why it is the default:

```{r}
#| label: both-tests
foresty_interaction(fit_no2, "no2", "sex", test = "both")
```

The likelihood ratio test is computed from the likelihood, so it has no robust
form, and a quasi-likelihood family and a GEE have no likelihood at all. Where
it cannot be taken, or where robust standard errors would leave it out of step
with the intervals beside it, the Wald test is reported in its place and
`foresty` says so; ask for one by name and you get that one or an error.

A label can be written where the variable is named, which saves repeating it in
`labels`:

```{r}
#| eval: false
foresty_interaction(fit_no2, exposure = c(NO2 = "no2"),
                    interaction = c(Sex = "sex"))
```

`foresty_combine()` puts figures already made together, which is how the
subgroup panel of a paper is built: the overall estimate at the top, one
subgroup analysis under another, and the interaction p-value beside each block.

```{r}
#| label: combine
#| fig-height: 3.6
overall <- foresty_main(list(fit_no2), exposure = "no2",
                        labels = c(no2 = "NO2"))
by_smoking <- foresty_interaction(fit_no2, "no2", "maternal_smoking")

foresty_combine(
  Overall = overall,
  Sex = by_sex,
  `Maternal smoking` = by_smoking
)
```

Nothing is refitted and nothing is re-estimated; every row is the row it was
drawn from on its own figure. A named argument names its block, and an unnamed
one is named for itself — the modifier for a `foresty_interaction()` figure,
`"Overall"` for a `foresty_main()` one. A block holding a single row is drawn
as that row rather than as a heading with one row under it. The figures have to
agree on the effect measure and the confidence level, since the rows are read
against one axis; they do not have to come from the same model, and usually
will not.

The title names the measure, the outcome it is a measure of and the exposure
the figure reports, and it names the exposure once however many blocks it was
drawn under: a figure whose overall block called it `NO2` and whose subgroup
blocks called it `no2` is titled with the first of those, not with both.

### The overall estimate

The overall estimate is not one of the subgroups — it is what they are read
against — so it is drawn apart from them: a filled diamond on its confidence
interval, half again the size of the squares under it, its label in bold, and a
rule between it and the subgroups. The interval is a plain line, as every other
interval on the figure is, so the whole figure is read the same way and the
diamond says which row is the summary.

`emphasize` says which blocks are drawn that way, and `emphasize = NULL` draws
every block alike:

```{r}
#| eval: false
foresty_combine(Overall = overall, Sex = by_sex, emphasize = NULL)
```

`emphasis_shape`, `emphasis_size`, `emphasis_face` and `emphasis_gap` in the
layout say how they are drawn. `emphasis_shape = "diamond"` draws instead the
wide summary diamond a paper draws, spanning the confidence interval in place
of it, at `emphasis_height`.

### Several exposures are several figures

Every row of a combined figure is read against the rows above it, and rows
reporting the effect of different exposures are not comparable that way,
however alike their axes happen to be. So `foresty_combine()` sorts the figures
it is given by the exposure they report and draws one figure for each — NO2 on
the first, black carbon on the second — each titled with its own exposure:

```{r}
#| label: combine-two
#| fig-height: 2.6
by_sex_bc <- foresty_interaction(fit_bc, "black_carbon", "sex")

figures <- foresty_combine(
  Overall = foresty_main(list(fit_no2, fit_bc),
                         exposure = c(NO2 = "no2",
                                      `Black carbon` = "black_carbon")),
  `NO2 by sex` = by_sex,
  `Black carbon by sex` = by_sex_bc
)

names(figures)
figures
```

A call covering one exposure — the usual one — returns that figure, as it
always did. A call covering several returns them in a list, named for the
exposures, which draws them one after another when it is printed and holds
`foresty` figures that are used singly as any other is:

```{r}
#| eval: false
figures[["NO2"]]
ggplot2::ggsave("no2.png", figures[["NO2"]], width = 9, height = 3)
```

## The HTML report

`foresty_report()` writes a self-contained page holding the model, the
estimates, the interaction test, the ratio of the odds ratios between levels,
the whole coefficient table and the figure. Nothing is linked from outside it,
so it can be sent to a co-author as it is.

```{r}
#| eval: false
foresty_report(by_sex, file = "no2_sex.html")
```

Nothing is written unless it is asked for. `html = TRUE` writes the same page
as the figure is made, under a name written from the variables it is about —
the exposure and the modifier, joined by an underscore — so a directory of them
can be read without opening any:

```{r}
#| eval: false
foresty_interaction(fit_no2, "no2", "sex", html = TRUE)   # -> no2_sex.html
foresty_main(list(fit_no2), "no2", html = TRUE)           # -> no2.html
foresty_main(list(fit_no2, fit_bc), c("no2", "black_carbon"),
             html = TRUE)                                 # -> no2_black_carbon.html
```

A path is taken as it is, as `html = "reports/no2_sex.html"`, and `FALSE`, the
default, writes nothing. A page for a figure drawn from several models carries
each of them, in a section of its own headed with the exposure it was fitted
for; `model = 1` cuts that down to one.

The report needs `gt`, which is not installed with the package.

## The result is a ggplot

Layers, scales and themes are added to it as usual, and `+` always reaches the
forest:

```{r}
#| eval: false
by_sex + ggplot2::coord_cartesian(xlim = c(0.9, 1.2))
by_sex + ggplot2::labs(title = "Infancy NO2 and childhood asthma")
by_sex + ggplot2::theme_minimal(base_size = 15)
```

The estimates travel with the figure, so `summary()`, `tidy()` and the rest
still work after it has been retuned.

`&` reaches every panel of a figure that has more than one. That is worth
knowing about but rarely what you want, since a scale or a coordinate system
applied to the table of numbers will spoil it. Use `+`.

## The table beside the plot

The numbers are drawn next to the figure — the estimate with its interval, the
counts, and the p-value for the interaction where there is one — since a forest
plot is read from them as much as from the marks. `table = FALSE` leaves a
plain figure, and `columns` chooses what the table carries:

```{r}
#| eval: false
foresty_main(list(fit_no2, fit_bc), c("no2", "black_carbon"), table = FALSE)
by_sex_table <- foresty_interaction(fit_no2, "no2", "sex",
                                    columns = c("estimate", "n", "events"))
```

A figure reporting a test of an interaction leaves the p-value of each row off:
two columns of p-values side by side are read for each other however they are
headed, and the one a reader reaches for is the test of the interaction. Name
`"p"` in `columns` to have the other back.

The figure is a row of panels — the labels, the plot, the numbers — drawn on
one shared scale, so a number sits beside the interval it belongs to whatever
else is on the figure. Each column of text is as wide as the widest thing in
it and no wider, and the plot takes what they leave, which is what keeps a
crowded figure compact. A subgroup is named on a row of its own above the rows
it covers, and those rows are indented under it.

## Layout and style

`layout` says how the figure is drawn. Pass the name of a style:

```{r}
#| label: jama
foresty_interaction(fit_no2, "no2", "sex", layout = "jama")
```

| Style | What it does |
|---|---|
| `"classic"` | the default: black on white, a dashed line at the null, a rule under the headings |
| `"jama"` | numbers to the left of the plot, rules around the block, navy squares, p-values as `.03` and `<.001`, `No.` and `P value` as headings |
| `"nejm"` | numbers to the right, blue squares, en dashes between the confidence limits |
| `"lancet"` | the Lancet's blue, en dashes, and the middle dot for a decimal point (`1·17`) |
| `"bmj"` | the BMJ's violet, grey rules, `to` between the confidence limits |
| `"revman"` | Cochrane's RevMan: blue squares, intervals written `1.15 [0.93, 1.42]` |

They are approximations of a house style, read off the forest plots those
journals print and the layouts of the
[`meta`](https://github.com/guido-s/meta) package, not the style sheets
themselves. `foresty_layout()` starts from one and changes whatever it needs
to:

```{r}
#| eval: false
foresty_layout("jama", colour = "#B24745", base_size = 11)
foresty_layout("classic", palette = c(estimate = "#00468B", null = "grey60"))
```

Some of what it holds:

| | |
|---|---|
| `colour`, `palette` | one colour for the estimates and their intervals, or each colour on the figure by name |
| `base_size`, `family` | font size and family |
| `point_shape`, `point_size`, `interval_width`, `null_line` | the marks |
| `emphasis_shape`, `emphasis_height`, `emphasis_size`, `emphasis_face`, `emphasis_gap` | how the overall estimate is singled out on a combined figure |
| `rules`, `separators`, `band`, `grid`, `axis_line` | the lines and the shading |
| `group_position`, `table_side` | the subgroup on a row or in a column; the numbers to the left of the plot or to its right |
| `digits`, `p_format`, `decimal_mark`, `ci_separator`, `ci_brackets` | how the numbers are written |
| `headings` | the column headings, as `c(n = "No. of patients")` |
| `column_gap` | space between one column of numbers and the next, in ems |
| `xlim`, `arrows` | limits for the plot, and labels for the two directions |
| `plot_width` | how much of the figure the plot takes |

### How wide the plot is

Each column of text takes the centimetres its longest entry needs, and the plot
takes everything left over, so widening the figure widens the plot and nothing
else:

```{r}
#| eval: false
ggplot2::ggsave("figure.png", by_sex_table, width = 13, height = 3.2)
```

Three things make the columns give the plot more room. `headings` replaces the
estimate's heading, which is written from the model and is usually the widest
thing on the figure; `column_gap` tightens the space between the columns; and
`columns` drops the ones the figure does not need.

```{r}
#| label: tightened
foresty_interaction(
  fit_no2, "no2", "sex",
  columns = c("estimate", "interaction_p"),
  layout = foresty_layout("classic", headings = c(estimate = "aOR (95% CI)"),
                          column_gap = 0.6)
)
```

`plot_width` decides it outright. A fraction gives the plot that share of the
figure at whatever width it is drawn — the columns of text then share the rest
between them, so give the figure enough width that they still fit:

```{r}
#| eval: false
foresty_layout("classic", plot_width = 0.6)   # three fifths of the figure
foresty_layout("classic", plot_width = 9)     # exactly 9 cm
```

`xlim` is worth knowing about: an interval running past the limits is drawn
with an arrow at that end rather than being left off, which keeps one wide
interval from flattening the rest of the figure.

```{r}
#| label: urbanicity
fit_urban <- glm(asthma ~ urbanicity + sex + maternal_age,
                 family = binomial, data = foresty_cohort)

foresty_main(
  list(fit_urban), "urbanicity",
  layout = foresty_layout(
    "classic",
    xlim = c(0.8, 2.5),
    arrows = c("Lower risk of asthma", "Higher risk of asthma"),
    headings = c(estimate = "aOR (95% CI)")
  )
)
```

The styles that follow a journal leave the title the functions write for
themselves off the figure, since that text belongs in the caption; a title
passed by hand is always drawn, and `title = NA` draws none whatever the style.

`headings` also names the column of row labels, which is headed `Exposure` by
`foresty_main()`, by the modifier by `foresty_interaction()` and `Subgroup` by
`foresty_combine()` — left unheaded there when the figure carries an overall
estimate, which is not a subgroup:

```{r}
#| eval: false
foresty_main(list(fit_no2, fit_bc), c("no2", "black_carbon"),
             layout = foresty_layout("classic",
                                     headings = c(label = "Pollutant")))
```

## Passing the result on

| | |
|---|---|
| `summary(x)` | the whole coefficient table with standard errors and p-values, laid out as `summary.glm()` lays one out, then the estimates and the interaction test |
| `as.data.frame(x)` | everything the figure carries, display labels included |
| `broom::tidy(x)` | the estimates in `broom`'s columns and order — `term`, `estimate`, `std.error`, `statistic`, `p.value`, `conf.low`, `conf.high`, then the counts. `tidy(x, what = "coefficients")` gives the model's coefficients instead |
| `broom::glance(x)` | one row: measure, n, events, person-time, interaction test |
| `predict(x, ...)` | passes through to the underlying model |
| `coef()`, `vcov()`, `formula()`, `nobs()`, `model.frame()` | likewise |

`tidy()` and `glance()` are `broom`'s generics, and `foresty` registers its
methods on them rather than carrying `broom` itself. Install `broom` if you
want them — it is not installed with the package — and call them as
`broom::tidy(x)` or after `library(broom)`. `as.data.frame()` needs nothing.

## Robust standard errors

`vcov = "robust"` gives the heteroskedasticity-consistent sandwich estimator
(`HC1`); `"HC0"` to `"HC4"` name one exactly. `cluster` gives the
cluster-robust version. Both come from `sandwich`.

```{r}
#| eval: false
foresty_main(list(fit_no2), "no2", vcov = "robust")
foresty_main(list(fit_no2), "no2", cluster = "practice_id")
```

`cluster` says which observations belong together, so it takes a column name, a
vector of one identifier per observation, or a one-sided formula — not `TRUE`.

Where sandwich does not apply, foresty says so rather than returning a
plausible wrong number: refit a Cox model with `coxph(..., robust = TRUE)`, and
an `rms` fit through `rms::robcov()`. A fit that is already robust is used as
it stands.

## What it does with the model

The subgroup effects come from **one** model containing the interaction, not
from a model refitted inside each subgroup. In the reference level of the
modifier the effect is the exposure term alone; in the other levels it is the
exposure term plus the relevant interaction terms. Every subgroup therefore
shares one estimate of the covariate effects, and the interaction can be
tested, which is what fitting separate models inside each subgroup gives up.

The linear combinations and their tests are computed by
[`car`](https://cran.r-project.org/package=car), and not reimplemented here.

The contrast is built as the difference between two rows of the model's own
design matrix, one at the baseline value of the exposure and one at the value
compared with it. Every column not involving the exposure cancels. That is why
a factor exposure, a factor with more than two levels, a spline among the
covariates, and `rms`'s own naming all come out right without the package
having to parse a single coefficient name.

## Models it works with

Anything supplying `coef()`, `vcov()` and a model frame. Tested against:

| | |
|---|---|
| base R | `glm()` (binomial, Poisson, Gaussian, any link), `lm()` |
| `survival` | `coxph()`, `survreg()` |
| `rms` | `lrm()`, `cph()`, `ols()`, `Glm()`, `psm()` |
| `lme4` | `lmer()`, `glmer()` — the fixed effects, with the random effects held as they were fitted |
| others | `MASS::polr()`, `nnet::multinom()`, `geepack::geeglm()` |

The effect measure is read from the model and named on the figure, and can be
overridden with `measure`:

| Model | Measure |
|---|---|
| logistic, `lrm()`, `polr()` | Odds ratio |
| Cox, `cph()` | Hazard ratio |
| log-link binomial | Risk ratio |
| Poisson with an offset | Incidence rate ratio |
| linear, `ols()` | Mean difference |

The measure names the outcome it is a measure of, so the axis under the plot
reads *Adjusted odds ratio for asthma* and the title *Adjusted odds ratio for
asthma associated with NO2*: what a ratio is a ratio of is the outcome, and the
exposure it is reported for comes after it. The outcome comes from the left of
the formula; a survival outcome is named by its event rather than its time. The
column of estimates beside the plot is headed with the measure alone, the axis
having named the outcome once already.

A ratio can be left on the scale the model was fitted on:

```{r}
#| eval: false
foresty_main(list(fit_no2), "no2", exponentiate = FALSE)
```

The figure then reports a log odds ratio read against zero rather than against
one, and `tidy()` returns the estimates and intervals on that scale too. It has
no effect on a mean difference or a coefficient, which are on that scale
already.

The axis is linear. A log axis would put a halving and a doubling the same
distance from the null, which is worth having when estimates span several fold,
but it is not the default — add `+ ggplot2::scale_x_continuous(transform =
"log")` when you want it.

Linear models are tested with *t* and *F* on their residual degrees of freedom;
everything else with the normal approximation and a Wald chi-square. The
interaction is tested by the likelihood ratio, or by the Wald test, or by both,
through `test` in `foresty_interaction()`; a quasi-likelihood family and a GEE
have no likelihood to take the first one over, and a mixed model fitted by REML
is refitted by maximum likelihood before it is taken, as `lme4`'s own `anova()`
does.

A model carrying time — a Cox model, or a Poisson model with an offset —
reports the person-time behind each row alongside the events, since a rate
cannot be read without it:

```{r}
#| label: rate
rate <- glm(asthma ~ no2 + sex + maternal_smoking + maternal_age +
              offset(log(followup_years)),
            family = poisson, data = foresty_cohort)

foresty_main(list(rate), exposure = "no2", labels = c(no2 = "NO2, ug/m3"),
             contrast = 10)
```

The offset makes the measure an incidence rate ratio rather than a risk ratio,
the figure reports it per 10 µg/m³ and says so on the row, and the column of
person-time is drawn beside the counts. The subgroup and combined figures
carry it the same way:

```{r}
#| eval: false
foresty_combine(
  Overall = foresty_main(list(rate), exposure = c(NO2 = "no2"), contrast = 10),
  Sex = foresty_interaction(rate, exposure = "no2", interaction = "sex",
                            contrast = 10)
)
```

## What it will not do

* **Cross two modifiers with each other.** The package reports two-way
  interactions: one exposure, one modifier, one joint test. Two modifiers are
  two calls and two blocks of a combined figure, which is what a subgroup panel
  is; a three-way interaction is not fitted for you.
* **Split a continuous modifier for you.** Categorize it with `cut()` first and
  refit. Which cut points you use is a decision about the analysis, not a
  default the package should pick.
* **Summarise a splined exposure with one number.** Its effect depends on where
  along the spline the difference is taken, so `at = c(from, to)` has to name
  the two values. The basis has to be in the formula — `splines::ns(no2, 3)`,
  `rms::rcs(no2, knots)`, `poly()` — because the two rows of the design matrix
  are built by evaluating it at those values. A basis computed beforehand and
  fitted as `y ~ spline1 + spline2 + spline3` records nothing that ties those
  columns to the variable, so naming the variable is an error that says as
  much; refit with the basis in the formula, which is the same model.
* **Invent a unit.** A continuous exposure is labelled with its name, and with
  the increment where that is not one — `contrast = 10` draws the row as
  `NO2 (per 10)` — but never with a unit, because the package cannot know what
  the numbers in a column mean. Name the unit in `labels`, as
  `c(no2 = "NO2, ug/m3")`, and the row reads `NO2, ug/m3 (per 10)`.
* **Draw an exposure that is interacted with something, from `foresty_main()`.**
  It has no single effect in such a model, and the function says so rather than
  reporting one of them.

## Dependencies

`car`, `checkmate`, `ggplot2`, `patchwork`, `scales`, `grid`, and base R.

Everything else is installed only if you want what it does: `broom` for
`tidy()` and `glance()`, `gt` for the HTML report, `sandwich` for robust
standard errors, `tibble` so that `tidy()` returns one, and `lme4`, `rms` and
`survival` if you fit models with them.

## License

GPL-3.
